\documentclass[11pt]{article}

\input{../mathsnotes.tex}

% Haim Brezis numbers the lemmas separately
\cleartheorem{lemma}
\theoremstyle{plain}
\newtheorem{lemma}{Lemma}[section]

\begin{document}
\begin{center}
\textbf{\large Theorems from Brezis' \textit{Functional Analysis, Sobolev Spaces and Partial Differential Equations} (first edition)} \\
Gustaf Bjurstam\\
bjurstam@kth.se\\
\end{center}

\section{The Hahn-Banach Theorems. Introduction to Conjugate Convex Functions}
\begin{theorem}[Hahn-Banach, analytic]
    Let $E$ be a vector space over $\real$, and let $p:E\to\real$ be a Minkowski functional. Let $G$ be a linear subspace of $E$ and let $g:G\to \real$ be a linear functional such that $g(x)\leq p(x)$ for all $x\in G$. There exists a linear functional $f:E\to \real$ such that
    \begin{equation*}
        f(x)=g(x),\, \forall x\in G,
    \end{equation*}
    and 
    \begin{equation*}
        f(x)\leq p(x),\, \forall x\in E.
    \end{equation*}
\end{theorem}

\begin{lemma}[Zorn]
    Every nonempty ordered set that is inductive has a maximal element.
\end{lemma}

\begin{corollary}
    Let $G$ be a linear subspace of the real vector space $E$. If $g\in G^\star$, then there exists $f\in E^\star$ that extends $g$ and such that $\norm{f}_{E^\star} = \norm{g}_{G^\star}$.
\end{corollary}

\begin{corollary}
    Let $E$ be a normed real vector space. For every $x\in E$ there is $f\in E^\star$ such that $\norm{f} = \norm{x}$ and $\inner{f}{x} = \norm{x}^2$.
\end{corollary}

\begin{corollary}
    For every $x$ in the real normed vector space $E$ we have
    \begin{equation*}
        \norm{x} = \sup_{\substack{f\in E^\star \\ \norm{f}\leq 1}} \abs{\inner{f}{x}} = \max_{\substack{f\in E^\star \\ \norm{f}\leq 1}} \abs{\inner{f}{x}}.
    \end{equation*}
\end{corollary}

\begin{proposition}
    Let $E$ be a real normed vector space, and let $H=[f=\alpha]\subset E$ be an affine hyperplane. Then $H$ is closed if and only $f$ is continuous.
\end{proposition}

\begin{theorem}[Hahn-Banach, 1st geometric]
    Let $E$ be a normed vector space, and let $A,B\subset E$ be two nonempty convex subsets such that $A\cap B = \empty$. Assume $A$ is open, then there exists a closed hyperplane $H$ that separates $A$ and $B$.
\end{theorem}

\begin{lemma}
    Let $E$ be a normed vector space, and let $C\subset E$ be an open convex set with $0\in C$. For every $x\in E$ set $p(x)= \inf \set{\alpha : \alpha^{-1}x\in C}$. Then $p$ is a Minkowski functional. Furthermore, there is a constant $M$ such that $0\leq p(x)\leq M\norm{x}$, and $C = \set{x\in E: p(x)<1}$.
\end{lemma}

\begin{lemma}
    Let $E$ be a normed vector space, and let $C\subset E$ be an open convex set. Assume $x_0\in E\setminus C$. Then there exists $f\in E^\star$ such that $f(x)< f(x_0)$ for all $x\in C$. In particular, $[f=f(x_0)]$ separates $C$ and $\set{x_0}$.
\end{lemma}

\end{document}
